HO unification from object language to meta language

### 2024-02-15 main.tex
rephrase tex

\noindent In our type-class solver, we compile the \coqIn{forall_dec} instance
of \coq (on the left) into the \elpi rule on the right. This compilation, which
is blindly built from the type of \coqIn{forall_dec}, hides some unification
properties accepted by the unification algorithm of the \ol but that are
rejected by the unification algorithm of the \ml. The main problem of this
compilation is how the subterm \coqIn{~$\forall$~x, P x} is translated in
the \ml rule; \coqIn{P} is a higher order variable, in the former subterm, it
does see the binder \coqIn{x}. However, the rule in the \ml, consider a first
order variable \elpiIn{P}, not seeing \elpiIn{x}. Furthermore, this subterm is
lodged within the \elpiIn{app} constructor making unification very structural.

@@

\noindent The two example of queries above, already in the syntax of the \ml,
represent two terms that would unify in the \ol, but fail in the \ml. In
particular, the first fails since the \ml is not able to unify \elpiIn{const
`nat} with the term \elpiIn{app[P, x]}: they have different rigid heads.

@@

Following the definition give just before, the terms \elpiIn{app[const `f, const
`y, x]} and \elpiIn{app[P, x]} should unify with substitution $\theta := \{P
\mapsto \texttt{\elpiIn{app [const `f, const `y]}} \}$. However this unification
is not possible in the \ml since, this time we have two terms with the same head
(the \elpiIn{app} node), but the corresponding lists have different lengths.

An immediate solution to solve these unification problems\todo{mettere da qualche parte anche eta e beta}
would be to be lazy and build very general rule for each instance in the database.
This would build a rule with the following shape:

@@

\noindent In this second implementation, any call to the \elpiIn{tc-Decision}
will unify with the head of the rule for \coqIn{forall_dec}. The the first
premise will ensure that the arguments unify with the what we expect from the
instance definition. Notice that, this time, the very unification algorithm of
the \ol is exploited, overcoming the unification problems listed just before.
This approach is quite appealing but it has two main disadvantages: the first
lays on the fact that the communication between the \ol and the \ml can
potentially be slow, this is due to the transformation of terms from one language
to the other. \todo{here i refer to constr->elpi e elpi->constr}
The other drawback of this latter encoding is that clause discrimination cannot
be done on the head of the clause, for instance, any indexing 
algorithm allowing to filter rules become suddenly useless.

>>>>

\noindent In our type-class solver, we translate the \coqIn{forall_dec} instance
of \coq (on the left) into the \elpi rule (on the right). This translation, derived
directly from the type of \coqIn{forall_dec}, conceals certain unification
properties that are accepted by the unification algorithm of the \ol but 
rejected by the unification algorithm of the \ml. The primary issue of this
translation lies in how the subterm \coqIn{~$\forall$~x, P x} is rendered in
the \ml rule; \coqIn{P} is a higher-order variable, and in the original subterm, it
sees the binder \coqIn{x}. However, the rule in the \ml interprets \elpiIn{P} as a 
first-order variable, not seeing \elpiIn{x}. Additionally, this subterm is
nested within the \elpiIn{app} constructor, introducing a structural challenge 
to the unification process.

@@

\noindent The two examples above, already expressed in the syntax of the \ml,
depict two terms that would successfully unify in the \ol, but encounter failure
in the \ml. Specifically, the first query fails because the \ml is not able to unify \elpiIn{const
`nat} with the term \elpiIn{app[P, x]} due to their distinct rigid heads.

@@

Based on the definition provided earlier, the terms \elpiIn{app[const `f, const
`y, x]} and \elpiIn{app[P, x]} are expected to unify with substitution $\theta := \{P
\mapsto \texttt{\elpiIn{app [const `f, const `y]}} \}$. However this unification
is not possible in the \ml. This difficulty arises because, in this instance,
we have two terms with the same head
(the \elpiIn{app} node), but their corresponding lists have different lengths.

% Implementation using ol unification on each argument
An immediate approach to address these unification problems\todo{mettere da qualche parte anche eta e beta}
would be to adopt a lazy strategy and construct highly general rules for each instance in the database.
This approach would result in a rule with the following structure:

@@

\noindent In this second implementation, any call to the \elpiIn{tc-Decision}
will unify with the head of the rule for \coqIn{forall_dec}. The first
premise ensures that the arguments unify with what we expect from the
instance definition. Notably, in this case, the unification algorithm of
the \ol is leveraged, , successfully addressing the previously mentioned unification problems.

However, this approach has two main drawbacks. Firstly, the communication
between the \ol and the \ml may potentially be slow due to the transformation of
terms from one language to the other \todo{here I refer to constr->elpi and
elpi->constr}. Secondly, the alternative encoding lacks the ability to
discriminate clauses based on the head of the clause. For example, any indexing
algorithm designed to filter rules becomes ineffective in this scenario.
